`{peacesciencer}`: An R Package for Quantitative Peace Science Research

### 2021-11-16 user-guide/ms.Rmd
update

`{peacesciencer}` comes with a fully developed suite of built-in functions for generating some of the most widespread forms of peace science data and populating the data with important variables that recur in many quantitative analyses. The core functionality of `{peacesciencer}` reduces to two broad categories of functions. These categories are functions that create the base data of interest to a researcher and functions, called after the base data are created, that add variables of interest to the data frame or subset the base data to a handful of rows that the researcher deems appropriate for analysis. Table \@ref(tab:tab-createfunctions) and Table \@ref(tab:tab-addfunctions) list these core functions as of version 0.7, the version of this package slated for release alongside the manuscript.[^helperfunctions]

@@

Table \@ref(tab:tab-createfunctions) are functions that create base data frames for a researcher starting an original project. They serve as functions that communicate the units of analysis supported in this package and that the package is capable of generating for an interested user. For example, `create_stateyears()` will generate the full universe of state-years from the Correlates of War [@cowstates2016, v. 2016] or Gleditsch-Ward [@gledtischward1999rlis, v.2017] system, encompassing all state-years to the most recently concluded calendar year, depending on the arguments supplied to the user in the function. `create_leaderyears()` will generate the full universe of leader-years from the Archigos leader data [@goemansetal2009ia, v. 4.1], optionally standardizing leader-years to the Gleditsch-Ward or Correlates of War state system data. As of version 0.7, `{peacesciencer}` is capable of creating full dyad-year data, leader-day data, leader-dyad-year data, leader-year data, state-day data, and state-year data.[^leaderdyadyear] A vignette on the package's website shows how users can create other forms of data from these functions as well (e.g. dyadic-dispute-year, leader-months, state-quarters).

@@

[^whittle]: The "whittle" (`wc_`) class of functions are a suite of functions that whittle dyadic dispute-years to true dyad-year data. They are effectively used in `add_cow_mids()` and `add_gml_mids()` for dyad-year data, but are offered here to allow users to employ their own case exclusion rules to create their own dyad-year conflict data. A vignette on `{peacesciencer}`'s website explains these case exclusion rules in some detail.

@@

[^poundsign]: R users will recognize that pound signs (`#`) allow for commenting code. Users new to R should understand its use here is primarily for exposition of what the code snippets in this manuscript are doing.

@@

I encourage users who are using `{peacesciencer}` for the first time, especially those who are learning R for the first time because of their interest in this package, to approach `{peacesciencer}` with an idea of the kind of data they want to create for the sake of a project. The core functionality of `{peacesciencer}` begins with the creation of a data frame, after which can be populated with various indicators of interest. No matter, the suggested use of `{peacesciencer}` begins with the creation of a data frame. To start, assume a new user without much familiarity with the R programming language installed RStudio, `{tidyverse}`, and `{peacesciencer}` with the idea of using `{peacesciencer}` to help them start a new research project that seeks to explain civil conflict onset across state-years. Toward that end, they have identified their unit of analysis is state-year and can be created with the `create_stateyears()`. To get started, I encourage entering the following command in the console in RStudio.

@@

A user who approaches `{peacesciencer}` with a project in mind will see they can better tailor this function to what they want. Their interest in a state-year analysis of civil conflict will likely gravitate them toward the Gleditsch-Ward state system data, since that is the state system that serves as the basis of the UCDP armed conflict data. They will also know they have no use for pre-1946 observations since civil conflict data, certainly UCDP's suite of data, have coverage only from 1946 forward. Thus, the user can supply some additional arguments to tailor the creation of data to just what they want (here: all Gleditsch-Ward state-years from 1946 to 2019).

@@

Another reader may be interested in using `{peacesciencer}` for a research project using a leader-year unit of analysis, which can be created with `create_leaderyears()`. Users can read more about what this function is doing by consulting the documentation file in R with the following command.

@@

Adding information to this data frame is a simple matter of joining a series of functions together in a "pipe." The "pipe"---represented as `%>%` in the code below---is an operator built into `{tidyverse}` that allows (`R`) users to pass forward expressions or functions. These pipes are common in the programming world and, as the code below will show, have the benefit of changing code in a way that is more intuitive and easier to both read and write for the user. While these functions can be modified to work without `{tidyverse}` installed and loaded into the session, the user will find their experience with `{peacesciencer}` is only improved by this important package.

@@

Finally, we can add some covariates of interest to these data. `add_peace_years()` calculates peace spells between ongoing conflicts in the data generated by `add_ucdp_acd()`. `add_democracy()` adds information about the level of democracy in the year using three prominent data sets on democracy (Polity, V-Dem, and Marquez' [-@marquez2016qme] extension of Pemstein et al's [-@pemsteinetal2010dc] Unified Democracy Scores). `add_creg_fractionalization()` adds information about the fractionalization and polarization of a state's ethnic and religious groups from the Composition of Religious and Ethnic Groups (CREG) Project at the University of Illinois. `add_sdp_gdp()` will add information about a state's estimated GDP, population, and GDP per capita from the @andersetal2020bbgb simulations. Finally, `add_rugged_terrain()` provide two estimates of the ruggedness of a state's terrain. The first is the terrain ruggedness index calculated by @nunnpuga2012r and the second is the @giblermiller2014etts extension of the natural logged percentage of the state that is mountainous (originally calculated by @fearonlaitin2003eicw). At the end of the pipe, the data returned by `{peacesciencer}` is assigned to an object minimally called `Data`.

@@

Table \@ref(tab:tab-addfunctions) shows there are some dedicated functions for populating leader-level data with leader-specific information, in addition to adding state-year-level information (e.g. democracy, capabilities) to leader-level data.  Table \@ref(tab:tab-createfunctions) shows support for creating leader-level data that are standardized to either the CoW or G-W state system data. Suppose a researcher wanted to create leader-year data, standardized to CoW system dates, for an analysis of inter-state dispute initiation analogous to what @horowitzstam2014hpme do in their analysis. The user would first start with creating the base data (`create_leaderyears(standardize = "cow", subset_years=c(1875:2010))`), which also generates some leader attributes (e.g. estimated leader age, year in office, and leader gender). From there, they would add information about leader conflict behavior in the year with `add_gml_mids()` and calculate peace-years with the `add_spells()` function.[^oninitiation] `add_lead()` will add a battery of leader experience and attributes variables from data created by @ellisetal2015lead, including whether the leader had military experience, combat experience, was a rebel fighter, and many more. Finally, `add_nmc()` and `add_democracy()` will add state-year estimates of national capabilities and democracy for assisting in making state-to-state comparisons, even for leader-level analyses. Some light recoding of the data created in `{peacesciencer}` and some regression modeling reproduces the results presented in Table \@ref(tab:tab-ly), itself an approximation of the kind of leader-year analysis exemplified in @horowitzstam2014hpme.

@@

NewGene is the latest evolution of EUGene, at least as a stand-alone executable program for creating the kind of data of interest to the peace science community. NewGene's greatest strength, relative to EUGene and even `{peacesciencer}`, is its support for *k*-adic data [c.f. @poast2010mdd]. *k*-adic levels of analysis are not yet supported in `{peacesciencer}` and users interested in generating *k*-adic data should consider downloading NewGene. No matter, `{peacesciencer}` has several superlatives in relation to NewGene. It supports leader-year and leader-dyad-year data while NewGene does not. It offers support for the G-W (and UCDP) ecosystem of data whereas NewGene does not and is mostly aimed for researchers interested in inter-state conflict. NewGene deviates a bit from EUGene by only indirectly asking users what kind of data they want (e.g. state-year, dyad-year) and without providing too much detail about what it is doing and why it is doing what it is doing. For example, NewGene only indirectly states the unit of analysis of the data to be generated near the top of its interface by asking the user how many country columns they want. This is an indirect way of the user to get data that are state-year, dyad-year, triad-year (etc.), which are then expanded and populated with data at various levels. `{peacesciencer}`, much like EUGene, is more explicit, encouraging the user to be upfront about what their unit of analysis is and what are the units of analysis that can be plausibly plugged into the data the user is creating. `{peacesciencer}`, again, does expect at least some level of comfort with the R programming language, but even this comes with greater ease of interpreting what is the unit of analysis and what are the primary spatial and temporal units that serve as the basis of the data.

>>>>

`{peacesciencer}` comes with a fully developed suite of built-in functions for generating some of the most widespread forms of peace science data and populating the data with important variables that recur in many quantitative analyses. The core functionality of `{peacesciencer}` reduces to two broad categories of functions. These categories are functions that create the base data of interest to a researcher and functions, called after the base data are created, that add variables of interest to the data frame or subset the base data to a handful of rows that the researcher deems appropriate for analysis.[^leftjoin] Table \@ref(tab:tab-createfunctions) and Table \@ref(tab:tab-addfunctions) list these core functions as of version 0.7, the version of this package slated for release alongside the manuscript.[^helperfunctions]

[^leftjoin]: In both cases, SQL-like "joins"---especially the left (outer) join---are the primary workhorse in these functions. A vignette on the package website, included in the appendix, explains these various joins, why they are used, and what issues may arise from them.

@@

Table \@ref(tab:tab-createfunctions) lists functions that create base data frames for a researcher starting an original project. They serve as functions that communicate the units of analysis supported in this package and that the package is capable of generating for an interested user. For example, `create_stateyears()` will generate the full universe of state-years from the Correlates of War [@cowstates2016, v. 2016] or Gleditsch-Ward [@gledtischward1999rlis, v.2017] system, encompassing all state-years to the most recently concluded calendar year, depending on the arguments supplied to the user in the function. `create_leaderyears()` will generate the full universe of leader-years from the Archigos leader data [@goemansetal2009ia, v. 4.1], optionally standardizing leader-years to the Gleditsch-Ward or Correlates of War state system data. As of version 0.7, `{peacesciencer}` is capable of creating full dyad-year data, leader-day data, leader-dyad-year data, leader-year data, state-day data, and state-year data.[^leaderdyadyear] A vignette on the package's website shows how users can create other forms of data from these functions as well (e.g. dyadic-dispute-year, leader-months, state-quarters).

@@

[^whittle]: The "whittle" (`wc_`) class of functions is a suite of functions that whittle dyadic dispute-years to true dyad-year data. They are effectively used in `add_cow_mids()` and `add_gml_mids()` for dyad-year data, but are offered here to allow users to employ their own case exclusion rules to create their own dyad-year conflict data. A vignette on `{peacesciencer}`'s website explains these case exclusion rules in some detail.

@@

[^poundsign]: R users will recognize that pound signs (`#`) allow for commenting code. Users new to R should understand its use here is primarily for exposition of what the code snippets in this manuscript are doing. In subsequent code snippets, `#>` will indicate console output the user can expect from these functions.

@@

I encourage users who are using `{peacesciencer}` for the first time, especially those who are learning R for the first time because of their interest in this package, to approach `{peacesciencer}` with an idea of the kind of data they want to create for the sake of a project. The core functionality of `{peacesciencer}` begins with the creation of a data frame, which can then be populated with various indicators of interest. No matter, the suggested use of `{peacesciencer}` begins with the creation of a data frame. To start, assume a new user without much familiarity with the R programming language installed RStudio, `{tidyverse}`, and `{peacesciencer}` with the idea of using `{peacesciencer}` to help them start a new research project that seeks to explain civil conflict onset across state-years. Toward that end, they have identified their unit of analysis is state-year and can be created with the `create_stateyears()`. To get started, I encourage entering the following command in the console in RStudio.

@@

A user who approaches `{peacesciencer}` with a project in mind will see they can better tailor this function to what they want. Their interest in a state-year analysis of civil conflict will likely gravitate them toward the Gleditsch-Ward state system data, since that is the state system that serves as the basis of the UCDP armed conflict data. They will also know they have no use for pre-1946 observations since civil conflict data typically have coverage only from 1946 forward. Thus, the user can supply some additional arguments to tailor the creation of data to just what they want (here: all Gleditsch-Ward state-years from 1946 to 2019).

@@

Another reader may be interested in using `{peacesciencer}` for a research project using a leader-year unit of analysis. These data can be created with `create_leaderyears()`. Users can read more about what this function is doing by consulting the documentation file in R with the following command.

@@

Adding information to this data frame is a simple matter of joining a series of functions together in a "pipe." The "pipe"---represented as `%>%` in the code below---is an operator built into `{tidyverse}` that allows users to pass forward expressions or functions. These pipes are common in the programming world and, as the code below will show, have the benefit of changing code in a way that is more intuitive and easier to both read and write for the user. While these functions can be modified to work without `{tidyverse}` installed and loaded into the session, the user will find their experience with `{peacesciencer}` is only improved by this important package.

@@

Finally, we can add some covariates of interest to these data. `add_spells()` calculates peace spells between ongoing conflicts in the data generated by `add_ucdp_acd()`. `add_democracy()` adds information about the level of democracy in the year using three prominent data sets on democracy (Polity, V-Dem, and Marquez' [-@marquez2016qme] extension of Pemstein et al's [-@pemsteinetal2010dc] Unified Democracy Scores). `add_creg_fractionalization()` adds information about the fractionalization and polarization of a state's ethnic and religious groups from the Composition of Religious and Ethnic Groups (CREG) Project at the University of Illinois. `add_sdp_gdp()` will add information about a state's estimated GDP, population, and GDP per capita from the @andersetal2020bbgb simulations. Finally, `add_rugged_terrain()` provide two estimates of the ruggedness of a state's terrain. The first is the terrain ruggedness index calculated by @nunnpuga2012r and the second is the @giblermiller2014etts extension of the natural logged percentage of the state that is mountainous (originally calculated by @fearonlaitin2003eicw). At the end of the pipe, the data returned by `{peacesciencer}` is assigned to an object minimally called `Data`.

@@

Table \@ref(tab:tab-addfunctions) shows there are some dedicated functions for populating leader-level data with leader-specific information, in addition to adding state-year-level information (e.g. democracy, capabilities) to leader-level data.  Table \@ref(tab:tab-createfunctions) shows support for creating leader-level data that are standardized to either the CoW or G-W state system data. Suppose a researcher wanted to create leader-year data, standardized to CoW system dates, for an analysis of inter-state dispute initiation analogous to what @horowitzstam2014hpme do. The user would first start with creating the base data (`create_leaderyears(standardize = "cow", subset_years=c(1875:2010))`), which also generates some leader attributes (e.g. estimated leader age, year in office, and leader gender). From there, they would add information about leader conflict behavior in the year with `add_gml_mids()` and calculate peace-years with the `add_spells()` function.[^oninitiation] `add_lead()` will add a battery of leader experience and attributes variables from data created by @ellisetal2015lead, including whether the leader had military experience, combat experience, was a rebel fighter, and many more. Finally, `add_nmc()` and `add_democracy()` will add state-year estimates of national capabilities and democracy for assisting in making state-to-state comparisons, even for leader-level analyses. Some light recoding of the data created in `{peacesciencer}` and some regression modeling reproduces the results presented in Table \@ref(tab:tab-ly), itself an approximation of the kind of leader-year analysis exemplified in @horowitzstam2014hpme.

@@

NewGene is the latest evolution of EUGene, at least as a stand-alone executable program for creating the kind of data of interest to the peace science community. NewGene's greatest strength, relative to EUGene and even `{peacesciencer}`, is its support for *k*-adic data [c.f. @poast2010mdd]. *k*-adic levels of analysis are not yet supported in `{peacesciencer}` and users interested in generating *k*-adic data should consider downloading NewGene. No matter, `{peacesciencer}` has several superlatives in relation to NewGene. It supports leader-year and leader-dyad-year data while NewGene does not. It offers support for the G-W (and UCDP) ecosystem of data whereas NewGene does not and is mostly aimed for researchers interested in inter-state conflict. NewGene deviates a bit from EUGene by only indirectly asking users what kind of data they want (e.g. state-year, dyad-year) and without providing too much detail about what it is doing and why it is doing what it is doing. For example, NewGene only indirectly states the unit of analysis of the data to be generated near the top of its interface by asking the user how many country columns they want. This is an indirect way of the user to get data that are state-year, dyad-year, triad-year (etc.), which are then expanded and populated with data at various levels. `{peacesciencer}`, much like EUGene, is more explicit, encouraging the user to be upfront about what their unit of analysis is and what are the data that can be plausibly plugged into the data the user is creating. `{peacesciencer}`, again, does expect at least some level of comfort with the R programming language, but even this comes with greater ease of interpreting what is the unit of analysis and what are the primary spatial and temporal units that serve as the basis of the data.
